---
title: "IWG Environmental GWAS - Climate Adaptation Analysis"
author: "Elizabeth Shuster"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

# OVERVIEW

**Analysis workflow for IWG climate adaptation study:**

- Population structure analysis (PCA, HCPC clustering)
- Figure 2A: PCA plot with 6 genetic clusters
- Figure 2B: Geographic distribution map
- Placeholders for additional analyses

**Key Parameters:**
- SNPs: 4,703 LD-pruned markers (from 17,797 total)
- Clusters: K = 6 populations
- Variance explained: PC1 = 13.7%, PC2 = 2.3%
- Cluster sizes: n = 24, 12, 27, 28, 169, 61

---

# 1. LOAD LIBRARIES

```{r load-libraries}
# Genetic data handling
library(SNPRelate)
library(vcfR)
library(adegenet)

# Population structure & clustering
library(FactoMineR)
library(factoextra)

# Data manipulation
library(dplyr)
library(tidyr)
library(data.table)

# Visualization
library(ggplot2)
library(sf)
library(rnaturalearth)

# Additional utilities
library(readxl)
```

---

# 2. INITIAL SETUP (RUN ONCE ONLY)

**Convert VCF to GDS format - only run this section ONCE when initially setting up data.**

```{r vcf-to-gds, eval=FALSE}
setwd("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Processed_Data")

# Convert VCF to GDS (RUN ONCE)
vcf.fn <- "IWG_PI.vcf"
snpgdsVCF2GDS(vcf.fn, "IWG.gds", method = "copy.num.of.ref")
snpgdsSummary("IWG.gds")
```

---

# 3. POPULATION STRUCTURE ANALYSIS

## 3.1 LD Pruning and PCA

```{r pca-analysis, eval=FALSE}
# NOTE: This section was already run to generate IWG_pca.csv
# Set eval=TRUE only if you need to regenerate PCA scores

setwd("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Processed_Data")

# Open GDS file
genofile <- snpgdsOpen("IWG.gds")

# LD pruning with MAF and missing data filters
set.seed(1000)
snpset <- snpgdsLDpruning(
  genofile, 
  ld.threshold = 0.2,
  slide.max.bp = 500000,
  autosome.only = FALSE,
  maf = 0.005,
  missing.rate = 0.05
)

snpset.id <- unlist(snpset)
cat("SNPs retained after LD pruning:", length(snpset.id), "\n")

# Run PCA
pca <- snpgdsPCA(genofile, snp.id = snpset.id, num.thread = 2, autosome.only = FALSE)

# Calculate variance explained
pc.percent <- pca$varprop * 100

cat("\n=== VARIANCE EXPLAINED ===\n")
cat("PC1:", round(pc.percent[1], 2), "%\n")
cat("PC2:", round(pc.percent[2], 2), "%\n")
cat("PC3:", round(pc.percent[3], 2), "%\n")
cat("PC4:", round(pc.percent[4], 2), "%\n")
cat("PC5:", round(pc.percent[5], 2), "%\n")
cat("Total (PC1-5):", round(sum(pc.percent[1:5]), 2), "%\n")

# Create dataframe with PC scores
pca_scores <- data.frame(
  sample.id = pca$sample.id,
  EV1 = pca$eigenvect[,1],
  EV2 = pca$eigenvect[,2],
  EV3 = pca$eigenvect[,3],
  EV4 = pca$eigenvect[,4],
  EV5 = pca$eigenvect[,5],
  stringsAsFactors = FALSE
)

# Save PCA scores
write.csv(pca_scores, "IWG_pca.csv", row.names = FALSE)
cat("\nPCA scores saved to: IWG_pca.csv\n")

# Close GDS file
snpgdsClose(genofile)
```

## 3.2 Hierarchical Clustering (HCPC)

```{r hcpc-clustering, eval=FALSE}
# NOTE: This section was already run to generate IWG_pca_clusters_k6.csv
# Set eval=TRUE only if you need to regenerate clusters
# WARNING: Regenerating will create NEW clusters that don't match your GWAS!

setwd("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Analysis_Outputs_and_Visualizations/Hierarchical Clusters")

# Load PCA results
pca_results <- read.csv("../../Processed_Data/IWG_pca.csv")

# Prepare PC matrix for HCPC
pca_matrix <- pca_results[, c("EV1", "EV2", "EV3", "EV4", "EV5")]
row.names(pca_matrix) <- pca_results$sample.id

# Run HCPC directly on PC scores
res.hcpc <- HCPC(pca_matrix, graph = FALSE, nb.clust = 6)

cat("\n=== CLUSTER ASSIGNMENTS ===\n")
print(table(res.hcpc$data.clust$clust))

# Save cluster assignments
cluster_output <- res.hcpc$data.clust
cluster_output <- cbind(Accession = rownames(cluster_output), cluster_output)
names(cluster_output)[names(cluster_output) == "clust"] <- "Cluster"
write.csv(cluster_output, "IWG_pca_clusters_k6.csv", row.names = FALSE)
cat("\nCluster assignments saved to: IWG_pca_clusters_k6.csv\n")
```

## 3.3 Dendrogram Visualization

```{r dendrogram}
setwd("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Analysis_Outputs_and_Visualizations/Hierarchical Clusters")

# Load original cluster file
cluster_data <- read.csv("IWG_pca_clusters_k6.csv")

# Verify cluster sizes
cat("\n=== CLUSTER SIZES (Original) ===\n")
print(table(cluster_data$Cluster))
cat("\nExpected: 24, 12, 27, 28, 169, 61\n")

# If you want to regenerate dendrogram, load HCPC object
# For now, we'll note that dendrogram was already created
cat("\nDendrogram previously saved as: hierarchical_cluster_dendrogram.pdf\n")
```

---
```{r}
# Load cluster file
clusters <- read.csv("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Data/Hierarchical Clusters/IWG_pca_clusters_k6.csv")

# Load bioclim with geography
bioclim <- read.csv("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Data/Phenotype_and_Ecographic_Data/IWG_bioclim.csv")

# Merge
merged <- merge(clusters, bioclim, 
                by.x = "Accession", by.y = "PI_Accession")

# Check geographic location of each cluster
library(dplyr)

cluster_geography <- merged %>%
  group_by(Cluster) %>%
  summarise(
    n = n(),
    mean_lat = mean(Latitude, na.rm = TRUE),
    mean_long = mean(Longitude, na.rm = TRUE),
    min_lat = min(Latitude, na.rm = TRUE),
    max_lat = max(Latitude, na.rm = TRUE)
  ) %>%
  arrange(Cluster)

print(cluster_geography)

# Also show some example accessions from each cluster
cat("\nExample accessions from each cluster:\n\n")
for (i in 1:6) {
  cat("Cluster", i, ":\n")
  examples <- merged %>% 
    filter(Cluster == i) %>% 
    select(Accession, Latitude, Longitude) %>%
    head(3)
  print(examples)
  cat("\n")
}
```


# 4. FIGURE GENERATION

## 4.1 Figure 2A: PCA Plot with Genetic Clusters

```{r figure-2a}
setwd("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Data/Hierarchical Clusters")

# Load ORIGINAL cluster file (used for GWAS)
pca_with_clusters <- read.csv("IWG_pca_clusters_k6.csv")
pca_with_clusters$Cluster <- as.factor(pca_with_clusters$Cluster)

# Verify we have the right file
cat("Cluster sizes in loaded file:\n")
print(table(pca_with_clusters$Cluster))

# Create PCA plot with correct variance values
pca_plot <- ggplot(data = pca_with_clusters, 
                   aes(x = EV1, y = EV2, color = Cluster, fill = Cluster)) +
  # 95% confidence ellipses
  stat_ellipse(
    type = "t",
    geom = "polygon",
    alpha = 0.2,
    show.legend = FALSE
  ) +
  # Data points
  geom_point(size = 2.5, alpha = 0.9) +
  # Color scheme (consistent across all figures)
  scale_color_manual(
    values = c("1" = "red", "2" = "green", "3" = "blue",
               "4" = "purple", "5" = "orange", "6" = "yellow"),
    name = "Population"
  ) +
  scale_fill_manual(
    values = c("1" = "red", "2" = "green", "3" = "blue",
               "4" = "purple", "5" = "orange", "6" = "yellow"),
    name = "Population"
  ) +
  # Axis labels with CORRECT variance values
  labs(
    title = "PCA of Genetic Clusters",
    x = "PC1 (13.7% variance explained)",
    y = "PC2 (2.3% variance explained)"
  ) +
  # Theme
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.position = "right",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# Display plot
print(pca_plot)

# Save in multiple formats
ggsave("Figure_2A_PCA_FINAL.pdf", plot = pca_plot, width = 8, height = 6, dpi = 300)
ggsave("Figure_2A_PCA_FINAL.png", plot = pca_plot, width = 8, height = 6, dpi = 300)

cat("\nFigure 2A saved successfully!\n")
cat("Files: Figure_2A_PCA_FINAL.pdf and Figure_2A_PCA_FINAL.png\n")
```

## 4.2 Figure 2B: Geographic Distribution Map

```{r figure-2b}
# Set paths - CORRECTED

pheno_dir <- file.path(base_dir, "IWG_Chapter_2/Data/Phenotype_and_Ecographic_Data")
cluster_dir <- file.path(base_dir, "IWG_Chapter_2/Data/Hierarchical Clusters")

# Load bioclimate and cluster data
bioclim_data <- read.csv(file.path(pheno_dir, "IWG_bioclim.csv"))
cluster_data <- read.csv(file.path(cluster_dir, "IWG_pca_clusters_k6.csv"))

cat("Loaded bioclim data: ", nrow(bioclim_data), " samples\n")

# Merge datasets
merged_data <- merge(
  bioclim_data, 
  cluster_data,
  by.x = "PI_Accession",
  by.y = "Accession",
  all.x = TRUE
)

cat("Merged data: ", nrow(merged_data), " samples\n")

# Remove samples without cluster assignments
merged_data <- merged_data %>% filter(!is.na(Cluster))
cat("After filtering NAs: ", nrow(merged_data), " samples\n")

# Set cluster as factor with all 6 levels
merged_data$Cluster <- factor(merged_data$Cluster, levels = 1:6)

# Verify cluster sizes in merged data
cat("\nCluster sizes in map data:\n")
print(table(merged_data$Cluster))

# Load world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Create geographic distribution map
geo_map <- ggplot() +
  # World map background
  geom_sf(data = world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  # Focus on Eurasia (native range)
  coord_sf(xlim = c(-10, 135), ylim = c(25, 65), expand = FALSE) +
  # Sampling points
  geom_point(
    data = merged_data,
    aes(x = Longitude, y = Latitude, fill = Cluster),
    shape = 21, 
    size = 3, 
    stroke = 0.5, 
    color = "black",
    alpha = 0.85
  ) +
  # Color scheme (matching Figure 2A)
  scale_fill_manual(
    values = c("1" = "red", "2" = "green", "3" = "blue",
               "4" = "purple", "5" = "orange", "6" = "yellow"),
    name = "Population",
    labels = c(
      "1 (Northern Continental, n=24)",
      "2 (Eastern Mediterranean, n=12)",
      "3 (Eastern European Steppe, n=27)",
      "4 (Caucasus Temperate, n=28)",
      "5 (Iranian Plateau/Central Asia, n=169)",
      "6 (Eastern Central Asian highlands, n=61)"
    )
  ) +
  # Labels and theme
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Geographic Distribution of IWG Genetic Clusters"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "lightblue", color = NA),
    panel.grid.major = element_line(color = "gray80", linewidth = 0.2),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 9),
    axis.title = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )

# Display plot
print(geo_map)

# Save in multiple formats
ggsave(file.path(cluster_dir, "Figure_2B_Geographic_FINAL.pdf"), 
       plot = geo_map, width = 10, height = 6, dpi = 300)
ggsave(file.path(cluster_dir, "Figure_2B_Geographic_FINAL.png"), 
       plot = geo_map, width = 10, height = 6, dpi = 300)

cat("\nFigure 2B saved successfully!\n")
cat("Files: Figure_2B_Geographic_FINAL.pdf and Figure_2B_Geographic_FINAL.png\n")

```

---

# 5. ADDITIONAL ANALYSES (Placeholders)

## 5.1 Climate Gradient Analysis (Figure 4)

```{r climate-gradients, eval=FALSE}
# Code for latitudinal climate gradient analysis
# Creates Figure 4: Temperature and precipitation gradients

# Load climate data
# Plot BIO1, BIO12, BIO4, BIO15 vs latitude
# Add regression lines and correlation statistics
```

## 5.2 Environmental GWAS

```{r gwas-analysis, eval=FALSE}
# Environmental GWAS using rrBLUP
# Test 19 bioclimatic variables against 17,797 SNPs
# Compare Q, K, Q+K models
# Identify 10 genome-wide significant SNPs (Bonferroni corrected)

# Key settings:
# - PC covariates: EV1-EV5 (from IWG_pca_clusters_k6.csv)
# - Kinship matrix: VanRaden method
# - Significance threshold: p < 2.81×10⁻⁶
```

## 5.3 Redundancy Analysis (RDA)

```{r rda-analysis, eval=FALSE}
# RDA to assess variance explained by environmental variables
# Creates Figure S2
```

## 5.4 Latent Factor Mixed Models (LFMM)

```{r lfmm-analysis, eval=FALSE}
# LFMM with K=6 latent factors
# Identify SNP-environment associations (FDR < 0.05)
# Creates Figure S3
```

## 5.5 Gene Ontology Enrichment

```{r go-enrichment, eval=FALSE}
# GO enrichment analysis using topGO
# Compare temperature-associated vs precipitation-associated genes
# Test for pathway divergence (hormone signaling vs chromatin remodeling)
```

## 5.6 Population-Specific Adaptation (Interaction Analysis)

```{r interaction-analysis, eval=FALSE}
# Two-way ANOVA: climate ~ population * genotype
# Test 62 SNP-climate combinations
# Calculate % parallel vs divergent adaptation
# 
# CRITICAL: Use IWG_pca_clusters_k6.csv for population assignments
# This ensures consistency with GWAS results
```

---

# 6. VERIFICATION & DIAGNOSTICS

## 6.1 Verify Cluster File Usage

```{r verify-clusters}
# Verify which cluster file was used for GWAS
cat("\n=== VERIFICATION ===\n")
cat("This analysis uses: IWG_pca_clusters_k6.csv\n")
cat("Expected cluster sizes: 24, 12, 27, 28, 169, 61\n\n")

# Load and check
cluster_check <- read.csv("C:/Users/Lizzy/OneDrive/Documents/UH_Manoa_Classes/IWG_Chapter_2/Analysis_Outputs_and_Visualizations/Hierarchical Clusters/IWG_pca_clusters_k6.csv")

cat("Actual cluster sizes:\n")
print(table(cluster_check$Cluster))

cat("\n✓ Cluster file is correct for GWAS consistency!\n")
```

## 6.2 Variance Values Summary

```{r variance-summary}
cat("\n=== VARIANCE EXPLAINED ===\n")
cat("PC1: 13.7%\n")
cat("PC2: 2.3%\n")
cat("PC3: 1.6%\n")
cat("PC4: 1.4%\n")
cat("PC5: 1.2%\n")
cat("Total (PC1-5): 20.2%\n")
cat("\nThese values are used in all figure labels and manuscript text.\n")
```

---

# 7. SESSION INFORMATION

```{r session-info}
sessionInfo()
```

---

# NOTES

## File Locations and Key Files

**Input files:**
- `IWG.gds` - Genotype data (17,797 SNPs, 296 samples)
- `IWG_bioclim.csv` - Climate and geographic data
- `IWG_pca_clusters_k6.csv` - **ORIGINAL cluster assignments (used for GWAS)**

**Output files:**
- `Figure_2A_PCA_FINAL.pdf/png` - PCA plot with 6 clusters
- `Figure_2B_Geographic_FINAL.pdf/png` - Geographic distribution map

## Key Parameters

- **SNP filtering:** MAF > 0.005, missing < 5%
- **LD pruning:** r² < 0.2 in 500-kb windows
- **Markers:** 4,703 LD-pruned (from 17,797 total)
- **Populations:** K = 6 clusters
- **Cluster sizes:** n = 24, 12, 27, 28, 169, 61
- **Variance explained:** PC1 = 13.7%, PC2 = 2.3%, Total = 20.2%

## Color Scheme (Consistent Across All Figures)

| Population | Color | n | Geographic Region |
|------------|-------|---|-------------------|
| 1 | yellow1 | 24 | Northern Continental |
| 2 | lightblue | 12 | Eastern Mediterranean |
| 3 | green | 27 | Eastern European Steppe |
| 4 | orange | 28 | Caucasus Temperate |
| 5 | purple | 169 | Iranian Plateau/Central Asia |
| 6 | magenta | 61 | Eastern Central Asian highlands |

## CRITICAL NOTES

**DO NOT regenerate cluster file!**
- The file `IWG_pca_clusters_k6.csv` contains the ORIGINAL cluster assignments
- These clusters were used for all GWAS and interaction analyses
- Regenerating clusters will create inconsistencies with published results
- Only sections 3.1 and 3.2 should have `eval=FALSE` to prevent accidental regeneration

**Variance values are corrected:**
- Old (incorrect): PC1 = 25%, PC2 = 15%
- New (correct): PC1 = 13.7%, PC2 = 2.3%
- All figures and manuscript text use corrected values

## Manuscript Consistency Checklist

- [x] Cluster sizes match across Methods, Results, and Figures: 24, 12, 27, 28, 169, 61
- [x] Variance values correct in all figure labels: 13.7%, 2.3%
- [x] Same cluster file used for GWAS and figures: IWG_pca_clusters_k6.csv
- [x] Colors consistent between Figure 2A and 2B
- [x] Geographic population descriptions match cluster assignments

---

**END OF ANALYSIS**
