---
title: "LFMM"
output: pdf_document
date: "2025-09-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(lfmm)
library(vcfR)
library(dplyr)
library(ggplot2)
```

## Load in the Data

```{r}
# Read VCF file and extract genotypes
vcf <- read.vcfR("IWG_PI.vcf")
gt <- extract.gt(vcf, return.alleles = TRUE)

# Method 1: Using extract.gt with element parameter (recommended)
# This directly extracts genotypes as numeric matrix
gt_numeric <- extract.gt(vcf, element = "GT", as.numeric = TRUE)
```

# Load environemntal 
```{r}
env_data <- read.csv("IWG_bioclim.csv", header = TRUE, stringsAsFactors = FALSE)

cc<-env_data[,1:3]
which(!complete.cases(cc)) # indentify which accessions have lat/long coordinates

rows_to_remove<-c(1,   2,   3,   4,   5,  21,  39,  40,  42,  45,  46,  47,  48,  50,  
                  56,  57,  58,  59,  60,  62,  65,  70,  71,  81, 307, 318, 319, 320, 321, 325, 327, 328,                    329, 330, 331)
env_data_clean <- env_data[-rows_to_remove, ]
```

#pull the first variable latitude from the environemntal data
```{r}
Latitude_data <- as.matrix(env_data_clean[ , "Latitude", drop = FALSE])
Latitude_data[is.na(Latitude_data)] <- mean(Latitude_data, na.rm = TRUE)

```

#match genotype and environmental data
```{r}
# 1. Check dimensions first
cat("VCF samples:", ncol(gt_numeric), "\n")
cat("CSV rows:", nrow(env_data_clean), "\n")

# 2. Find common samples
vcf_samples <- colnames(gt_numeric)
csv_samples <- env_data_clean$PI_Accession  # adjust column name

common_samples <- intersect(vcf_samples, csv_samples)
cat("Common samples:", length(common_samples), "\n")

# 3. Subset both datasets to common samples
gt_matched <- gt_numeric[, colnames(gt_numeric) %in% common_samples]
csv_matched <- env_data_clean[env_data_clean$PI_Accession %in% common_samples, ]

# 4. Ensure same sample order
csv_matched <- csv_matched[match(colnames(gt_matched), csv_matched$PI_Accession), ]

```

# Impute missing values in genotype_data
```{r}
genotype_data<-gt_matched
genotype_data <- apply(genotype_data, 2, function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x))
genotype_data <- t(genotype_data)

```

# Run the LFMM
```{r}
#mod.lfmm <- lfmm_ridge(Y = genotype_data, X = Latitude_data, K = 1)
mod.lfmm <- lfmm_ridge(Y = genotype_data, X = Latitude_data, K = 6)
pv <- lfmm_test(Y = genotype_data, X = Latitude_data, lfmm = mod.lfmm)
lfmm_results <- data.frame(SNP = 1:ncol(genotype_data), p_value = pv$pvalue[, 1], log_p = -log10(pv$pvalue[, 1]))

```

# Extract position and chromosome info
```{r}
lfmm_results <- cbind(lfmm_results, chromosome = as.factor(vcf@fix[, "CHROM"]), pos = as.numeric(vcf@fix[, "POS"]))
```

# Define chromosome lengths based on reference for correct cumulative positioning
```{r}
chromosome_lengths <- c(
  "J01" = 548726614,
  "J02" = 694191331,
  "J03" = 664523050,
  "J04" = 541250370 ,
  "J05" = 602542912,
  "J06" = 566711382,
  "J07" = 686173349,
  "S01" = 422989602,
  "S02" = 507709987,
  "S03" = 493345210,
  "S04" = 377353484,
  "S05" = 549394082,
  "S06" = 408503727,
  "S07" = 482783349,
  "V01" = 574115619,
  "V02" = 618570211,
  "V03" = 526408016,
  "V04" = 440983052,
  "V05" = 591402192,
  "V06" = 462045100,
  "V07" = 761377749 
)
cumulative_starts <- c(0, cumsum(chromosome_lengths[-length(chromosome_lengths)]))
```

# Map chromosomes and compute cumulative positions
```{r}
chromosome_mapping <- setNames(1:21, names(chromosome_lengths))
lfmm_results <- lfmm_results %>%
  mutate(
    chromosome_numeric = chromosome_mapping[as.character(chromosome)],
    cumulative_pos = pos + cumulative_starts[chromosome_numeric]
  ) %>%
  arrange(chromosome_numeric, cumulative_pos)

# Define significance thresholds
significance_threshold <- -log10(0.001)  # 99th percentile threshold
bonferroni_threshold <- -log10(0.05 / nrow(lfmm_results))  # Bonferroni threshold
```

# plot
```{r}
#asthetics
library(RColorBrewer)
library(MetBrewer)

# Calculate midpoints for chromosome labels
midpoints <- cumulative_starts + chromosome_lengths / 2

# Define colors using the "VanGogh1" palette from MetBrewer
chromosome_colors <- met.brewer("VanGogh1", n = length(unique(lfmm_results$chromosome_numeric)))


# Define updated chromosome labels with prefixes
chromosome_labels <- paste0(c(1:21, "_", names(chromosome_mapping)))

# Plot with cumulative positions, 99th percentile line, Bonferroni threshold, and customized labels
# Define updated chromosome labels with prefixes
chromosome_labels <- paste0(c(1:21), "_", names(chromosome_mapping))

# Perform FDR adjustment on p-values
lfmm_results$fdr_p_value <- p.adjust(lfmm_results$p_value, method = "BH")

# Convert FDR-adjusted p-values to -log10 scale
lfmm_results$log_fdr_p <- -log10(lfmm_results$fdr_p_value)

# Define the FDR significance threshold 
fdr_threshold <- -log10(0.001)

# Plot with cumulative positions, 99th percentile line, FDR threshold, and customized labels
p <- ggplot(lfmm_results, aes(x = cumulative_pos, y = log_p, color = as.factor(chromosome_numeric))) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = bonferroni_threshold, color = "red", linetype = "dashed") +  # bonferonni
  geom_hline(yintercept = fdr_threshold, color = "purple", linetype = "dashed") +  # FDR threshold
  labs(
    title = "SNP Outliers for Latitude for All Groupings",
    x = "Cumulative Genomic Position",
    y = "-log10(p-value)",
    color = "Chromosome"
  ) +
  # Add lines to separate chromosomes
  geom_vline(xintercept = cumulative_starts[-1], color = "grey", linetype = "dotted", size = 0.5) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8),
    plot.title = element_text(size = 10)
  ) +
  scale_x_continuous(
    breaks = cumulative_starts + chromosome_lengths / 2,
    labels = chromosome_labels
  ) +
  scale_color_manual(values = chromosome_colors)

p
```

# loop through all environmental variables
```{r}
# Identify environmental columns (exclude non-environmental columns)
# Adjust these column names based on your actual CSV structure
exclude_cols <- c("PI_Accession", "Latitude", "Longitude", "Cultivar_Name", "ID")  # Add other non-environmental columns as needed
env_cols <- setdiff(colnames(csv_matched), exclude_cols)

# Initialize list to store results
all_lfmm_results <- list()

# Loop through all environmental variables
for (env_var in env_cols) {
  cat("Processing environmental variable:", env_var, "\n")
  
  # Extract current environmental variable
  env_matrix <- as.matrix(csv_matched[, env_var, drop = FALSE])
  
  # Handle missing values - impute with mean
  env_matrix[is.na(env_matrix)] <- mean(env_matrix, na.rm = TRUE)
  
  # Check if there's variation in the environmental variable
  if (var(env_matrix, na.rm = TRUE) == 0) {
    cat("Skipping", env_var, "- no variation in data\n")
    next
  }
  
  # Run LFMM
  tryCatch({
    # You can adjust K value as needed
    mod.lfmm <- lfmm_ridge(Y = genotype_data, X = env_matrix, K = 6)
    pv <- lfmm_test(Y = genotype_data, X = env_matrix, lfmm = mod.lfmm)
    
    # Store results with chromosome and position information
    lfmm_results <- data.frame(
      SNP = 1:ncol(genotype_data), 
      p_value = pv$pvalue[, 1], 
      log_p = -log10(pv$pvalue[, 1]),
      env_variable = env_var
    )
    
    # Add chromosome and position information
    lfmm_results <- cbind(lfmm_results, 
                         chromosome = as.factor(vcf@fix[, "CHROM"]), 
                         pos = as.numeric(vcf@fix[, "POS"]))
    
    # Add results to list
    all_lfmm_results[[env_var]] <- lfmm_results
    
    cat("Completed LFMM for", env_var, "\n")
    
  }, error = function(e) {
    cat("Error processing", env_var, ":", e$message, "\n")
  })
}

# Combine all results into one data frame
if (length(all_lfmm_results) > 0) {
  combined_results <- do.call(rbind, all_lfmm_results)
  
  # Add row names to identify the environmental variable
  combined_results$result_id <- rownames(combined_results)
  
  # Save results
  write.csv(combined_results, "lfmm_all_environmental_variables.csv", row.names = FALSE)
  
  cat("Analysis complete! Results saved to 'lfmm_all_environmental_variables.csv'\n")
  cat("Total environmental variables processed:", length(all_lfmm_results), "\n")
  
  # Print summary
  cat("\nSummary of results:\n")
  for (env_var in names(all_lfmm_results)) {
    n_significant <- sum(all_lfmm_results[[env_var]]$p_value < 0.05, na.rm = TRUE)
    cat(env_var, ": ", n_significant, " significant SNPs (p < 0.05)\n")
  }
  
} else {
  cat("No environmental variables were successfully processed.\n")
}

# Optional: Create individual result files for each environmental variable
if (length(all_lfmm_results) > 0) {
  for (env_var in names(all_lfmm_results)) {
    filename <- paste0("lfmm_", env_var, ".csv")
    write.csv(all_lfmm_results[[env_var]], filename, row.names = FALSE)
  }
  cat("Individual result files also created for each environmental variable.\n")
}
```

###joint plot
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(MetBrewer)

# Define chromosome lengths based on reference for correct cumulative positioning
chromosome_lengths <- c(
  "J01" = 548726614,
  "J02" = 694191331,
  "J03" = 664523050,
  "J04" = 541250370,
  "J05" = 602542912,
  "J06" = 566711382,
  "J07" = 686173349,
  "S01" = 422989602,
  "S02" = 507709987,
  "S03" = 493345210,
  "S04" = 377353484,
  "S05" = 549394082,
  "S06" = 408503727,
  "S07" = 482783349,
  "V01" = 574115619,
  "V02" = 618570211,
  "V03" = 526408016,
  "V04" = 440983052,
  "V05" = 591402192,
  "V06" = 462045100,
  "V07" = 761377749 
)

cumulative_starts <- c(0, cumsum(chromosome_lengths[-length(chromosome_lengths)]))

# Map chromosomes and compute cumulative positions
chromosome_mapping <- setNames(1:21, names(chromosome_lengths))

# Process the combined results (assuming you have combined_results from the previous script)
combined_results_processed <- combined_results %>%
  mutate(
    chromosome_numeric = chromosome_mapping[as.character(chromosome)],
    cumulative_pos = pos + cumulative_starts[chromosome_numeric]
  ) %>%
  arrange(env_variable, chromosome_numeric, cumulative_pos)

# Calculate significance thresholds for each environmental variable
combined_results_processed <- combined_results_processed %>%
  group_by(env_variable) %>%
  mutate(
    # Bonferroni threshold per environmental variable
    bonferroni_threshold = -log10(0.05 / n()),
    # FDR adjustment per environmental variable
    fdr_p_value = p.adjust(p_value, method = "BH"),
    log_fdr_p = -log10(fdr_p_value)
  ) %>%
  ungroup()

# Define general significance thresholds
general_significance_threshold <- -log10(0.001)  # 99th percentile threshold
fdr_threshold <- -log10(0.01)  # FDR threshold

# Calculate midpoints for chromosome labels
midpoints <- cumulative_starts + chromosome_lengths / 2

# Define colors using the "VanGogh1" palette from MetBrewer
chromosome_colors <- met.brewer("VanGogh1", n = length(unique(combined_results_processed$chromosome_numeric)))

# Define updated chromosome labels with prefixes
chromosome_labels <- paste0(c(1:21), "_", names(chromosome_mapping))

# Create the faceted Manhattan plot
p <- ggplot(combined_results_processed, aes(x = cumulative_pos, y = log_p, color = as.factor(chromosome_numeric))) +
  geom_point(alpha = 0.6, size = 0.5) +
  # Add significance threshold lines
  geom_hline(data = combined_results_processed %>% 
               group_by(env_variable) %>% 
               slice(1) %>% 
               ungroup(), 
             aes(yintercept = bonferroni_threshold), 
             color = "red", linetype = "dashed", alpha = 0.7) +
  geom_hline(yintercept = fdr_threshold, color = "purple", linetype = "dashed", alpha = 0.7) +
  # Add lines to separate chromosomes
  geom_vline(xintercept = cumulative_starts[-1], color = "grey", linetype = "dotted", size = 0.3, alpha = 0.5) +
  # Facet by environmental variable
  facet_wrap(~ env_variable, scales = "free_y", ncol = 3) +
  labs(
    title = "Manhattan Plots for All Environmental Variables",
    x = "Cumulative Genomic Position",
    y = "-log10(p-value)",
    color = "Chromosome",
    caption = "Red dashed line: Bonferroni threshold per variable; Purple dashed line: FDR threshold"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 12, hjust = 0.5),
    strip.text = element_text(size = 8, face = "bold"),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.spacing = unit(0.5, "lines")
  ) +
  scale_x_continuous(
    breaks = cumulative_starts + chromosome_lengths / 2,
    labels = chromosome_labels
  ) +
  scale_color_manual(values = chromosome_colors, guide = guide_legend(nrow = 2)) +
  # Adjust y-axis to better show the data
  coord_cartesian(expand = FALSE)

# Display the plot
p

# Optional: Save the plot with appropriate dimensions
ggsave("manhattan_plot_all_environmental_variables.png", 
       plot = p, 
       width = 16, 
       height = 12, 
       dpi = 300, 
       units = "in")

```

```{r}
# Print summary statistics for each environmental variable
cat("Summary of significant SNPs per environmental variable:\n")
summary_stats <- combined_results_processed %>%
  group_by(env_variable) %>%
  summarise(
    total_snps = n(),
    bonferroni_significant = sum(p_value < (0.05 / n()), na.rm = TRUE),
    fdr_significant = sum(fdr_p_value < 0.01, na.rm = TRUE),
    p_001_significant = sum(p_value < 0.001, na.rm = TRUE),
    mean_log_p = round(mean(log_p, na.rm = TRUE), 3),
    max_log_p = round(max(log_p, na.rm = TRUE), 3),
    .groups = 'drop'
  )

print(summary_stats)

#write.csv(summary_stats, "IWG_lfmm_summary.csv",row.names=F)

# Optional: Create a summary plot showing number of significant SNPs per environmental variable
summary_plot <- summary_stats %>%
  select(env_variable, bonferroni_significant, fdr_significant, p_001_significant) %>%
  tidyr::pivot_longer(cols = -env_variable, names_to = "threshold_type", values_to = "count") %>%
  mutate(threshold_type = case_when(
    threshold_type == "bonferroni_significant" ~ "Bonferroni",
    threshold_type == "fdr_significant" ~ "FDR < 0.01",
    threshold_type == "p_001_significant" ~ "p < 0.001"
  )) %>%
  ggplot(aes(x = env_variable, y = count, fill = threshold_type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(
    title = "Number of Significant SNPs by Environmental Variable",
    x = "Environmental Variable",
    y = "Number of Significant SNPs",
    fill = "Significance Threshold"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")

# Display summary plot
summary_plot

# Save summary plot
ggsave("significant_snps_summary.png", 
       plot = summary_plot, 
       width = 12, 
       height = 6, 
       dpi = 300, 
       units = "in")
```






